\documentclass[12pt]{article}
\usepackage[left=3cm, right=3cm, top=2cm]{geometry}

%\usepackage[light,condensed,math]{anttor}
\usepackage[T1]{fontenc}
\usepackage{amsmath,graphicx}
\usepackage{hyperref}

%%cmark, xmark
\usepackage{amssymb}% http://ctan.org/pkg/amssymb
\usepackage{pifont}% http://ctan.org/pkg/pifont
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\usepackage[table]{xcolor}% http://ctan.org/pkg/xcolor
\usepackage{tcolorbox}
\usepackage{float}
\usepackage{tikz}

\usetikzlibrary{arrows,automata,positioning}

%%cmark, xmark
\usepackage{amssymb}% http://ctan.org/pkg/amssymb
\usepackage{pifont}% http://ctan.org/pkg/pifont
% \newcommand{\cmark}{\ding{51}}%
% \newcommand{\xmark}{\ding{55}}%
%\usepackage[table]{xcolor}% http://ctan.org/pkg/xcolor
% \usepackage{tcolorbox}
\usepackage{float}

%
% The following macro is used to generate the header.
%
\newcommand{\lecture}[2]{
   \pagestyle{myheadings}
   \thispagestyle{plain}
   \newpage
   \noindent
   \begin{center}
   \framebox{
      \vbox{\vspace{2mm}
    \hbox to 6.28in { {\bf CS F351: Theory of Computation
	\hfill \textbf{$12^{th}$} Nov, 2022} }
       \vspace{4mm}
       \hbox to 6.28in { {\Large \hfill #1: #2  \hfill} }
       \vspace{2mm}
       \hbox to 6.28in { {\it \hfill Scribe: Piyush Mohite (2020A7PS0114G)} }
      \vspace{2mm}}
   }
   \end{center}
   \vspace*{4mm}
}
%

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{exercise}[theorem]{exercise}

\begin{document}
\lecture{29}{Pumping Lemma for Context Free Languages}

\section*{Pumping Lemma for CFLs}

For any CFL $ L $, $\exists$ $ k $ $\in$ $\mathbb{N}$, such that for any word $ w $ $\in$ $ L $ with $|w| > k$, there exists a partition w = uvxyz such that following holds:

\begin{itemize}
   \item$ |vy| > 0 $ ; i.e. the part to be pumped should be non empty.
   \item $uv^ixy^iz \in L$, $ \forall i \in \mathbb{N}$
   \item $ |vxy| \leq k$.
\end{itemize}

\textbf{Note:} Pumping Lemma is a necessary condition for a language to be Context-Free. It is not sufficient.
\\ \\
\textbf{Example:} $ L = \{ a^nb^nc^n \mid n \geq 0 \}$ is not a CFL, as it fails the pumping lemma.
\textbf{Proof:}
\begin{itemize}
   \item[ $-$ ] Assume, for a contradiction, that $ L $ is a CFL. Hence, Pumping Lemma for CFL should hold for $ L $, since it is a necessary condition for a language to be CFL.
   \item[ $-$ ] let G be a grammar for this language. WLOG, consider G to be in CNF. Assume for the size of G is bounded above by $ p $.
   \item[ $-$ ] Since we want to pick a long enough word, it should have length more than 2p, as G is in CNF. Let $ k = 2p + 1 $.
   \item[ $-$ ] Let $ w = a^kb^kc^k $ be the word in $ L $ such that $ |w| > 2p $ (i.e. $ |w| \geq k $).
   \item[ $-$ ] To arrive at a contradiction, we need to show that for any spilt of this word of the form $ uvxyz, \exists i $ such that $ uv^ixy^z \notin L $.
   \item[ $-$ ] Following cases of word split can arise:
   \item[ ] \begin{itemize}
         \item[] \textbf{Case I}: Any part to be pumped (i.e $v$ or $y$) is word made of single letter. In such case, that part can be pumped one more time so that we get a new word that does not lie in $L$. This is true for all combinations of $v,y$. An example in which $v$ is a word of all $a$'s is given below.

            \[ w = a^{j1}.a^{j2}.a^{j3}.a^{j4}.a^{(k-j1-j2-j3-j4)}b^kc^k \]

            \[w = u.v.x.y.z\]

            \textbf{The new word formed after pumping is:}

            \[w' = u.v^2.x.y^2.z = a^{(k+j2+j4)}b^kc^k\]

            Similarly, it can be proved that any word $ w $ whose partition $ v $ or $ y $ is word made with 1 letter will produce a word $ w' \notin L $.

         \item[] \textbf{Case II}:Any part to be pumped (i.e. $v$ or $y$) is word made of 2 letters. In such case,
            that part can be pumped one more time so that we get a new word that does not lie in $L$. This is true for all combinations of $v,y$. An example in which $v$ is a word of $a$'s and $b$'s is given below.

            \[ w = a^{j1}.a^{k - j1}b^{j2}.b^{j3}.b^{j4}.b^{k - j2 - j3 - j4}.c^k \]

            \[w = u.v.x.y.z\]

            \textbf{The new word formed after pumping is:}

            \[w' = u.v^2.x.y^2.z = a^kb^{j2}a{(j-j1)}b^kc^k \notin L\]

            Similarly, it can be proved that any word $ w $ whose partition $ v $ or $ y $ is word made with 2 letters will produce a word $ w' \notin L $.

         \item[] \textbf{Case III}: Any part to be pumped (i.e. $v$ or $y$) is word made of 3 letters. In such case, that part can be pumped one more time so that we get a new word that does not lie in $L$. This is true for all combinations of $v,y$. An example in which $y$ is a word of $a$'s, $b$'s
            and $c$'s is given below.

            \[ w = a^{j1}.a^{k - j1}b^{j2}.b^{j3}.b^{j4}.b^{k - j2 - j3 - j4}.c^k \]

            \[w = u.v.x.y.z\]

            \textbf{The new word formed after pumping is:}

            \[w' = u.v^2.x.y^2.z = a^{(k+j2)}b^kc^{j4}a^{(k-j1-j2-j3)}b^kc^k \notin L\]

            Similarly, it can be proved that any word $ w $ whose partition $ v $ or $ y $ is word made with 3 letters will produce a word $ w' \notin L $.

      \end{itemize}

   \item[ $-$ ] Hence for any spilt of this word of the form $ uvxyz$, $ \exists i $ such that $ uv^ixy^iz \notin L $.
   \item[ $-$ ] But, this is a contradiction, as we assumed that $ L $ is a CFL. Therefore, it can be concluded that our assumption was wrong.
   \item[ $-$ ] Hence, $ L $ is not a CFL. Hence proved.

\end{itemize}



\end{document}
